---
title: Parallelization in loo (current state)
---

## What is parallelization?

<!-- TODO: what is parallelization, why we would want it, differences in multprocessing/multithreading, etc., how to do in R, how does mirai work -->

### Serial `mirai` execution

Currently, parallelization in loo goes through [three paths](https://github.com/stan-dev/loo/blob/05a6bd20af83c919ba7a5572e3e0a50426c17a77/R/importance_sampling.R#L206) @loo:

1. Serial execution using `lapply()`
2. Forking through `parallel::mclapply()`
3. Fresh processes through `parallel::parLapply()`

The serial execution path is taken when the `cores` argument is equal to 1, and forking is only available on Linux and MacOS. Since `mirai` @mirai neatly subsumes paths 2 and 3, a question we had was what would the performance hit be if we also removed 1--that is, if a user requested serial execution, what do we lose by using `mirai` with a single core. 

```{r}
#| label: all-equal-helper
#| include: false
same_numeric_values <- function(a, b) {
  all.equal(
    a |> unlist(use.names = FALSE) |> unname(),
    b |> unlist(use.names = FALSE) |> unname(),
    check.attributes = FALSE,
    tolerance = sqrt(.Machine$double.eps)
  ) |>
    isTRUE()
}
```

```{r}
#| label: serial-mirai-benchmark
log_ratios <- -loo::example_loglik_matrix()
S <- nrow(log_ratios)
N <- ncol(log_ratios)
tail_len <- loo:::n_pareto(r_eff = rep(1, N), S = S)

x <- Map(
  \(i) {
    list(
      log_ratios_i = log_ratios[, i],
      tail_len_i = tail_len[i]
    )
  },
  seq_len(N)
)

work <- function(s) {
  loo:::do_psis_i(
    log_ratios_i = s$log_ratios_i,
    tail_len_i = s$tail_len_i
  )$pareto_k
}

mirai::daemons(1)

results <- bench::mark(
  vapply = vapply(x, work, numeric(1)),
  lapply = lapply(x, work),
  mirai = mirai::mirai_map(x, work)[],
  vapply_to_list = vapply(x, work, numeric(1)) |> as.list(),
  purrr_map = purrr::map_dbl(x, work),
  check = same_numeric_values,
  iterations = 200
)
```

Note that `mirai_map()` and `lapply()` results are lists. `same_numeric_values()`'s definition is elided from output but simply checks that results are identical without counting unlisting as part of the benchmark (unlike `vapply_to_list`).

```{r}
#| label: serial-mirai-results
results |> summary()
results |> summary(relative = TRUE)
```

NB: the `mem_alloc` column should not be interpreted as total memory usage for `mirai`: `bench::mark()` records R heap allocations via `utils::Rprofmem()`, which is not able to track daemons' memory usage [@bench; @R; @mirai].

`mirai_map()` with a single daemon has a median runtime of 23.3ms, roughly 2.2x slower than the fastest in-process approach. Of course, in absolute terms the difference is negligible, adding only about 10ms. The memory usage of all in-process approaches are relatively close to one another. `vapply()`'s small memory overhead may probably be attributed to the extra work it does validating types and lengths and such @R.

In short, it is probably a good idea to directly test the effects of using `mirai` execution in `loo`, but it is likely that we will find that we should keep the unique single core execution path.

## Where is parallelization currently in use?

* `function()`
    * short description
* `function()`
    * short description

## Where could we benefit from introducing parallelization? {#sec-parallelization-benefit}

To investigate this question we use two approaches:

1) Inspecting the code base and figuring out at which points parallelization might be helpful ("analytic search")
2) Running performance tests on the loo pipeline, investigating where the bottlenecks are, and finding out whether we can improve things with parallelization.

### Analytic search

For each function/place in code where you think parallelization might help, do:
* Introduction
    * short description of function
    * why do you think that parallelization can improve things?
* Experiment: 
    * implement parallelization
    * record time (define which time you are using) for parallel and unparallel version
    *  report results (how much speed up we can get?)
*  Short conclusion

### Performance testing

* short description of approach for running performance test
* run performance test
* report results
* identify bottlenecks
* provide first ideas whether any of the identified bottlenecks can benefit from parallelization

## Vectorization

When browsing the codebase to find places where parallelization could be helpful (see @sec-parallelization-benefit), we found two potential speedups--in these places, we may be able to benefit from _vectorizing_ functions instead of parallelizing a loop. Vectorization is advantageous in these spots as we would be rewriting R loops as optimized matrix operations [@advr, secs. 24.5 - 24.7].

> Matrix algebra is a general example of vectorisation. There loops are executed by highly tuned external libraries like BLAS. If you can figure out a way to use matrix algebra to solve your problem, you’ll often get a very fast solution.
> - Wickham [@advr, sec. 24.5]

Of course, for these potential improvements, we should first write tests for correctness, then write benchmarks in the PR.

The more important of the two is in [`pseudobma_weights()`](https://github.com/stan-dev/loo/blob/master/R/loo_model_weights.R#L334), when performing a Bayesian bootstrap. This function is the backend for `loo_model_weights()`, and is used to average models using psuedo-Bayesian model averaging, optionally using a Bayesian bootstrap (which is referred to as psudo-BMA+). This pseudo-BMA+ path is the one we care about as it has the elidable loop. We currently loop over Bayesian bootstrap iterations where it seems quite plausible to swap to matrix operations. This could potentially be a large speedup as the loop is large, being of length `BB_n`, which defaults to 1,000 replications. There don't appear to be any reallocations in the loop, so the speedup will probably be dominated by the efficiency of matrix ops [@advr, secs. 5.3.1, 24.6]. This is tracked in #XXX.
<!-- TODO: PR -->

Similarly, though less interesting, is a gradient calculation in [`stacking_weights()`](https://github.com/stan-dev/loo/blob/master/R/loo_model_weights.R#L278), just above the previous function. `stacking_weights()` is similarly used for combining models using stacking, i.e., fitting a linear model to weigh the contending models. The gradient function is used for solving for those weights. However, the speedup here is less interesting because we are looping over `K`, the number of models being compared, which is usually small. However, we may as well take the (potential) free performance. This is tracked in #XXX.
<!-- TODO: PR -->

These two loops seem like low hanging fruit and well worth the time to implement.

## Nested parallelization

When updating `loo` we wish to avoid "nested parallelization", i.e. attempting to compute something in parallel, when already in a parallel process. Take this nonfunctioning `mirai` code as an example:

```r
# parallel over xs
mirai_map(
  xs,
  # parallel over x
  function(x) mirai_map(x, mean)[]
)
```

We first are doing something parallely over values of `xs`, but then we try to work in parallel again for each value from `xs`! The main problem here is oversubscription: we spawn `n` outer workers, let say, then for each we spawn another `m` inner workers. In the package context though, this implementation would be far more complex, and would probably be hidden completely from users. This becomes a problem when a user sets `n`, the outer parallel workers, to # cores - 1[^1], as the actual number of parallel tasks could be as large as $n * m$. Oversubscription is bad because, in this context, we are mostly running CPU heavy tasks where each process takes time and needs its own core. Oversubscription is promising more parallelization than what the hardware can provide--note that parallelization usually has some [fixed startup costs as well](../experiments/mirai_overhead.md), so we would probably lose performance. In `loo` specifically, we must be careful when parallelizing functions to ensure we do not accidentally have nested parallelization. 

<!-- TODO: update codes -->
[^1]: This is a common heuristic for number of threads/daemons/workers/etc. to use because it leaves 1 core free to do normal computer things, lest the system become unresponsive. Note this doesn't account for memory usage at all--if you have 8n memory but send out n tasks which each use "1" memory, you will run out of memory. See [the note on memory](memory.md).

### Where do we currently find nested parallelization in `loo`?
<!-- TODO -->
* short description of function
* do you see any challenges with the identified functions?
* should nested parallelisation run parallel or sequential?
    * provide experiments to stress your statement

### Potential problems with packages that make use of `loo` such that nested parallelization can occur?
However, we know that other packages also call `loo` functions, which is another potential vector for nested parallelization. To this end, it seems like `brms` is the only (Stan) package where this could even be vaguely possible, but doing so seems to necessitate wilful parallelization not provided natively by `brms`. This merits further investigation, but it seems like we can (softly) table concerns regarding external nested parallelization for now.

<!-- TODO -->
Make a list of packages of which you have thought of:
* package name
* Any issues that you see?
* (make a short note even if you don't see issues)

## Appendix

```{r}
#| label: session-info
#| cache: false
sessioninfo::session_info(
  pkgs = c("loo", "mirai", "bench", "purrr"),
  info = c("platform", "packages"),
  dependencies = FALSE
)
```
