---
title: Serial `mirai` Execution
---

Currently, parallelization in loo goes through [three paths](https://github.com/stan-dev/loo/blob/05a6bd20af83c919ba7a5572e3e0a50426c17a77/R/importance_sampling.R#L206):

1. Serial execution using `lapply()`
2. Forking through `parallel::mclapply()`
3. Fresh processes through `parallel::parLapply()`

Serial execution is taken when the `cores` argument is equal to 1, and forking is only available on Linux and MacOS. Since `mirai` neatly replaces 2 and 3, a question we had was what would the performance hit be if we also removed 1--that is, if a user requested serial execution, what do we lose by using `mirai` with a single core. Here is a simple experiment we started with:

```{r}
#| include: false
same_numeric_values <- function(a, b) {
  all.equal(
    a |> unlist(use.names = FALSE) |> unname(),
    b |> unlist(use.names = FALSE) |> unname(),
    check.attributes = FALSE,
    tolerance = sqrt(.Machine$double.eps)
  ) |>
    isTRUE()
}
```

```{r}
x <- rep(0.05, 20)

work <- function(s) {
  Sys.sleep(s)
  sqrt(s)
}

mirai::daemons(1)

results <- bench::mark(
  vapply = vapply(x, work, numeric(1)),
  lapply = lapply(x, work),
  mirai = mirai::mirai_map(x, work)[],
  purrr_map = purrr::map_dbl(x, work),
  check = same_numeric_values,
  iterations = 20
)
```

Note that `mirai_map()` and `lapply()` results are lists. `same_numeric_values()`'s definition is elided from output but it exists to check that results are identical without counting unlisting as part of the benchmark.

```{r}
results |> summary()
results |> summary(relative = TRUE)
```

As expected, the times are all essentially identical; we are more interested in the memory usage. We can see that, unsurprisingly, `vapply()` allocates the least memory, by a long shot. `mirai_map()` is worse than `lapply()` by a good margin too, using about 50x the memory.
