## Glossary {#sec-glossary}

### Concepts

#### Serial

Tasks run one at a time, strictly one after another with no overlap.

#### Synchronous

The calling code blocks and waits for a result before continuing. Does not imply serial as tasks may run in parallel while the caller waits.
Example: Parallel sampling across four chains but before we can continue, we have to wait until all four chains are finished.

#### Asynchronous

The calling code moves on immediately after dispatching a task and can check back later for results. Enables non-blocking workflows.

#### Parallel

Multiple tasks run at the same time across separate workers. Can be combined with either synchronous or asynchronous calling patterns. 
Examples:

+ Synchronous parallel: running four chains in parallel but blocking until all are done.
+ Asynchronous parallel: dispatching multiple model fits to background processes. The main session fires off each model fit without blocking, so you could immediately dispatch the next model. Each background process then internally runs its four chains in parallel (sync). 

#### Forking

A child process is created by copying the parent R session at the OS level. The child inherits the full environment (packages, variables) via copy-on-write memory. Unix/macOS only.

#### PSOCK cluster

Worker processes started as completely fresh R sessions, communicating with the controller over sockets. No shared memory; packages and variables must be exported explicitly.

#### Daemon

A persistent background R process that stays alive between tasks, waiting to receive and execute work. Used by `mirai` as its worker model.

#### Persistent workers

A worker process that remains alive between tasks and across multiple calls, retaining its state and loaded packages. Avoids the startup cost of spawning a new process per job. Contrast with temporary workers (e.g., forked processes in `mclapply()`) that are created for a single call and exit when it finishes.

### Functions

#### `lapply()`

Base R serial map. Applies a function to each element of list-like input sequentially in the host session. No workers, no parallelization overhead; considered as the baseline.

#### `mclapply()` (Unix only)

Parallel map via forking. Workers inherit the session environment automatically. Falls back to `lapply()` (i.e., serial execution) on Windows.

#### `parLapply()` (all OS)

Parallel map over a PSOCK cluster. Requires explicit setup (`makeCluster`), export of variables, and teardown (`stopCluster`). Workers persist until stopped.

#### `daemons()`

Launches $n$ persistent background workers for the current `mirai` session. Workers stay alive and idle until reassigned or the session ends. Workers can run on the same machine, or can be remote resources, and roughly can be considered as background R processes. See [the documentation](https://mirai.r-lib.org/reference/daemons.html) for more details.

#### `mirai_map()`

Parallel asynchronous map that dispatches tasks to daemons. Returns a list of unresolved `mirai` objects immediately. Append `[]` to block and collect all results.

#### `[]`

The collect operator on a `mirai` result. Blocks until all dispatched tasks are complete and returns their values.

## How does `mirai` work? {#sec-mirai-internals}

By default, `mirai` uses a dispatcher which _daemons_ call in to[^1]. `daemons(4)` on a laptop spawns four background R processes and hands each one an IPC address to connect back on, whereas `daemons(url = host_url())` opens the listener and waits, leaving the user free to start `daemon()` processes wherever they like. Since the connection is always initiated by the daemon, only the host needs a routable address, and going from local cores to a compute node changes the URL and nothing else. This is what allows `mirai`, and thus `loo`, to be so flexible about where computations are run---the host process needs to know very little about each daemon. Tasks travel over those connections as serialized R objects, and daemons block. NNG hands over a result as soon as it lands, so an idle daemon costs nothing and there is no supervising loop running in R. This is what makes `mirai()` asynchronous: what it returns is a placeholder and `[]` (see the glossary) is the blocking mechanism to actually collect results.

[^1]: Note that `mirai` is driven by `nanonext` [@nanonext], which binds NNG (Nanomsg Next Gen), a C messaging library.

By default `daemons()` also starts a dispatcher, which queues tasks and hands each one off as a daemon frees up, giving first-in-first-out (FIFO) scheduling instead of the fixed round-robin assignment used when `dispatcher = FALSE` (which is also what `parLapply()` essentially does). This matters wherever task costs are uneven. The dispatcher is also what `stop_mirai()` and `.timeout` rely on to cancel a task that is already running, and it is where the `memory` argument blocks new tasks until the bytes queued for dispatch drop under a cap.

### Reproducibility with `mirai`

Most parallelized computations in `loo` are deterministic, but some paths evaluate user-supplied functions on workers, and those functions may consume random numbers. Changing the parallel backend can therefore change their RNG behavior. This section records how `mirai` handles seeding and reproducibility. `mirai` uses the L'Ecuyer PRNG.

Even if we set the same seed, we won't get the same results from `lapply()` and `mirai_map()`. Here, we can see that even if we use the same seed (`set.seed(0)` before each call), randomly generating a few numbers from an exponential distribution changes depending on whether we call `lapply()` or `mirai_map()`. More specifically, both maps are essentially doing `rexp(1), rexp(2), rexp(3), rexp(4)`.[^2] In this snippet we are using 4 daemons, so in the `mirai_map()` call, we are starting 4 RNG streams which are being derived from the main stream (see [in source](https://github.com/r-lib/mirai/blob/19ce00aeaf5cbc9d583a1d1409137692dee1d932/R/daemons.R#L626)).

[^2]: Somewhat orthogonal to this section, but note that `with(set.seed(0), lapply(1:4, rexp) |> unlist())` is equivalent to `with(set.seed(0), rexp(10))`.

```{r}
#| label: serial-vs-mirai
set.seed(0)
str(lapply(1:4, rexp))

# using with() just for scoping
with(
  {
    set.seed(0) # set seed in main process
    mirai::daemons(4) # start 4 background processes
  },
  # mirai_map() is an async map
  # `[]` blocks to collect results
  str(mirai::mirai_map(1:4, rexp)[])
)
```

Note that by default if `daemons()` is called without setting a seed, the RNG streams aren't reproducible (just like results in serial execution without setting a seed)---ensuring "statistical validity but not numerical reproducibility between runs" [@mirai-2.5.0]. If we rerun the exact same `mirai` code, we do get the same results:

```{r}
#| label: mirai-repeat
with(
  {
    set.seed(0)
    mirai::daemons(4)
  },
  str(mirai::mirai_map(1:4, rexp)[])
)
```

With 2 daemons, `mirai_map(1:4, rexp)` still creates four tasks: `rexp(1), rexp(2), rexp(3), rexp(4)`. The first two tasks run on the two available daemons (so they have the same results as before); later tasks are assigned as daemons become free. Since RNG streams are tied to daemons, changing the number of daemons changes which stream each task uses, so the output changes. A plain `set.seed(0)` is just seeding the host session and making each daemon derive its RNG state from host.

```{r}
#| label: mirai-daemon-count
with(
  {
    set.seed(0)
    mirai::daemons(2)
  },
  str(mirai::mirai_map(1:4, rexp)[])
)
```

Setting a seed manually when launching daemons gets reproducibility (stability over runs), "regardless of the number of daemons used." [@mirai-2.5.0]. Underneath the hood, `mirai` creates a new deterministic RNG stream per `mirai()` evaluation, not per daemon, which ensures that results are stable irrespective of daemon count.

```{r}
#| label: mirai-seed-stable
with(
  {
    mirai::daemons(4, seed = 0)
  },
  str(mirai::mirai_map(1:4, rexp)[])
)

with(
  {
    mirai::daemons(3, seed = 0)
  },
  str(mirai::mirai_map(1:4, rexp)[])
)
```

Note that `seed = 0` means something different in base R versus in `mirai`, that is the same seed doesn't automatically guarantee the same results between serial and parallel execution. In essence, reproducibility is the users' responsibility, akin to compute setups.

## Session Info

```{r}
#| label: session-info
#| cache: false
#| warning: false
#| echo: false
renv::dependencies(here::here("docs", "report.qmd"), quiet = TRUE)$Package |>
  unique() |>
  c("sessioninfo") |>
  setdiff("rmarkdown") |>
  sessioninfo::session_info(
    pkgs = _,
    info = c("platform", "packages"),
    dependencies = FALSE
  )
```
